F\# Data: \textnormal{Making structured data first-class citizens}

### 2015-10-31 icfp-2015/paper.tex
Switch to use variants

% TODO: Adapt XML example 
% TODO: Switch from sum types to variant types
% TODO: Discuss 'stability' of the inference (minor change in schema does not cause dramatic cahnges)
% TODO: Say that any variant is a subtype of any other variant 
% TODO: Say that anything is a subtype of variant
% TODO: Say that variant is a last-resort thing
% TODO: Double check that bottom/top are used correctly
% TODO: Explain (even more) that there is no least upper bound, which is why Theorem 3 is good enough
% TODO: Avoid "no other rule applies" in the formal system
% TODO: Practical evaluation (lots of downloads, lots of Github - how often people customize samples?)

@@

\subsection{Processing XML data, alternative}

@@

The grammar below defines \emph{structural type} $\sigma$. We distinguish between \emph{non-nullable types}
that always have a valid value (written as $\hat{\sigma}$) and \emph{nullable types} that encompass missing 
and \kvd{null} values (written as $\sigma$). We write $\nu$ for record field names and $\nu?$
for record names, which can be empty:
%
\begin{equation*}
\begin{array}{rcl}
 \hat{\sigma} &\narrow{=}& \nu?\; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \} \\[0.1em]
                &\narrow{|}& \ident{float} \lsep \ident{decimal} \lsep \ident{int} \lsep \ident{bit} \lsep \ident{bool} \lsep \ident{string} 
 \\[0.6em] 
       \sigma &\narrow{=}& ~\hat{\sigma}~ \lsep \hat{\sigma}\;\,\kvd{option} \lsep [\sigma] \\[0.1em]
              &\narrow{|}& \sigma_1 + \ldots + \sigma_n \lsep ~\top~ \lsep \kvd{null}
\end{array}
\end{equation*}
%
The non-nullable types include records (consisting of an optional name and zero or more fields with
their types) and primitive types. Records arising from XML documents are always named, while records
used by JSON and CSV providers are always unnamed.

@@

Finally, a union type in our model implicitly permits the \kvd{null} value. This is because structural union 
types are \emph{not} mapped to F\# union types. Instead our encoding requires the user to handle the 
cases one-by-one and so they also need to handle the situation when none of the cases matches (as discussed
in Section~\ref{sec:impl-unions}, this is the right choice for an F\# type provider, but it would not 
necessarily be a good fit for other languages).

% TODO: Also include reference to section with more on heterogeneous collections.

@@

\begin{definition}
We write $\sigma_1 :> \sigma_2$ to denote that $\sigma_2$ is a subtype of $\sigma_1$. The 
subtyping relation is defined as a transitive reflexive closure of these rules:

@@

\noindent\boxed{\emph{Union types}}
\begin{align}
\label{eq:sub-union1}\tag{U1}
\sigma_1 + \ldots + \sigma_n :> \sigma_1' + \ldots + \sigma_n' \quad &(\textnormal{iff}~\forall i.\sigma_i :> \sigma_i') \\
\label{eq:sub-union2}\tag{U2}
\sigma_1 + \ldots + \sigma_n :> \sigma_1 + \ldots + \sigma_m \quad &(\textnormal{where}~m \leq n) \\
\label{eq:sub-union3}\tag{U3}
\sigma_1 + \ldots + \sigma_n :> \sigma_1 + \ldots + \sigma_m \quad &(\textnormal{where}~m \geq n) \\
\label{eq:sub-union4}\tag{U4}
\hspace{-1em}\sigma_1 + \ldots + \sigma_n :> \sigma_{\pi(1)} + \ldots + \sigma_{\pi(n)} \quad &(\textnormal{permutation}~\pi) \\
\label{eq:sub-union-el}\tag{U5}
\sigma_1 + \ldots + \sigma_n :> \sigma_i \quad &(\textnormal{for}~i \in 1 \ldots n)
\end{align}

\noindent\boxed{\emph{Record types}}
\begin{align}
\hspace{-1em}
\label{eq:sub-record1}\tag{R1}
 \nu?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} \hspace{-0.1em}:> \hspace{-0.1em}
 \nu?~\{ \nu_1\!:\!\sigma_1', .., \nu_n\!:\!\sigma_n' \}
 \; (\sigma_i \hspace{-0.25em}:>\hspace{-0.25em} \sigma_i')
\end{align}
\vspace{-1.4em}
\begin{align}
\hspace{-1em}
\label{eq:sub-record2}\tag{R2}
 \nu?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} \hspace{-0.1em} :> \hspace{-0.1em}
 \nu?~\{ \nu_1\!:\!\sigma_1, .., \nu_m\!:\!\sigma_m \}
 \; (m \hspace{-0.1em}\geq\hspace{-0.1em} n)
\end{align}
\vspace{-1.0em}
\begin{align}
\label{eq:sub-record3}\tag{R3}
\hspace{-1.75em}
\begin{array}{l}
 \nu?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} :> \\
 \nu?~\{ \nu_{\pi(1)}\!:\!\sigma_{\pi(1)}, .., \nu_{\pi(m)}\!:\!\sigma_{\pi(m)} \}
\end{array} \qquad &(\textnormal{permutation}~\pi)
\end{align}
\vspace{-0.6em}
\begin{align}
\label{eq:sub-record4}\tag{R4}
\hspace{-1.75em}
\begin{array}{l}
 \nu?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n, \nu_{n+1}\!:\!\kvd{null}, .., \nu_{n+m}\!:\!\kvd{null} \} :> \\
 \nu?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \}
\end{array}
\end{align}
\end{definition}

@@

\item The \ident{bit} type is a subtype of both \ident{int} and \ident{bool} (\ref{eq:sub-prim}). 
  Thus, a sample $0,1$ has a type \ident{bit}; $0,1,\kvd{true}$ is \ident{bool} and
  $0,1,2$ becomes \ident{int} (for a sample $0,1,2, \kvd{true}$ we would need a union type).

@@

\item There is a top type (\ref{eq:sub-top}), but no bottom type. It is possible to find common 
  supertype of any two types, because a union type $\tau_1 + \ldots + \tau_n$ is a supertype of all its 
  components (\ref{eq:sub-union-el}).

\item For unions, we include the standard rules. Subtype can have fewer alternatives 
  (\ref{eq:sub-union1}), elements can be reordered (\ref{eq:sub-union2}) and subtyping is covariant 
  (\ref{eq:sub-union4}). An unusual rule (\ref{eq:sub-union3}) says that subtype can also have 
  \emph{more} cases. This is valid because the type provider exposes unions as objects that do not
  allow complete pattern matching.

@@

\noindent
The rule that allows subtype to have fewer record elements (\ref{eq:sub-record4}) is particularly
important. It allows us to prefer records in some cases. For example, given two samples 
$\{ \ident{name}:\ident{string} \}$ and $\{ \ident{name}:\ident{string}, \ident{age}:\ident{int} \}$,
we can find a common supertype $\{ \ident{name}:\ident{string}, \ident{age}:\ident{int}~\kvd{option} \}$
which is also a record. For usability reasons, we prefer this to another common supertype
$\{ \ident{name}:\ident{string} \} + \{ \ident{name}:\ident{string}, \ident{age}:\ident{int} \}$.
The next section describes precisely how our inference algorithm works.

@@

We could also be more or less strict about missing data -- we choose to handle missing values 
silently when we can (\kvd{null} becomes an empty collection), but we are explicit in other cases 
(we infer \ident{string}~\kvd{option} as a hint to the user rather than treating \kvd{null} as an 
empty string). These choices are based on experience with using F\# Data in practice.

@@

As mentioned earlier, the structural type inference relies on a finding common supertype. However,
the partially ordered set of types does not have a unique greatest lower bound. For example, record types
$\{\ident{a}:\ident{int}\}$ and $\{\ident{b}:\ident{bool}\}$ have common supertypes
$\{\ident{?a}:\ident{int}, \ident{?b}:\ident{bool}\}$ and $\{\ident{a}:\ident{int}\} + \{\ident{b}:\ident{bool}\}$
which are unrelated. Our inference algorithm always prefers records over unions (Theorem~\ref{thm:no-unions}). 
The definition of the \emph{common supertype} relation uses a number of auxiliary definitions 
that are explained in the discussion that follows.

>>>>

% TODO: Adapt XML example 
% TODO: Switch from sum types to variant types
% TODO: Discuss 'stability' of the inference (minor change in schema does not cause dramatic cahnges)
% TODO: Say that any variant is a subtype of any other variant 
% TODO: Say that anything is a subtype of variant
% TODO: Say that variant is a last-resort thing
% TODO: Double check that bottom/top are used correctly
% TODO: Explain (even more) that there is no least upper bound, which is why Theorem 3 is good enough
% TODO:  -- but maybe we can have lub w.r.t variants? or something?
% TODO: Avoid "no other rule applies" in the formal system
% TODO: Practical evaluation (lots of downloads, lots of Github - how often people customize samples?)

@@

\subsection{Processing XML data, alternative}
\label{sec:providers-xml}

@@

The grammar below defines \emph{structural type} $\sigma$. We distinguish between \emph{non-nullable types}
that always have a valid value (written as $\hat{\sigma}$) and \emph{nullable types} that encompass missing 
and \kvd{null} values (written as $\sigma$). We write $\nu$ for record field names and $\nu_?$
for record names, which can be empty:
%
\begin{equation*}
\begin{array}{rcl}
 \hat{\sigma} &\narrow{=}& \nu_?\; \{ \nu_1 : \sigma_1, \ldots, \nu_n : \sigma_n \} \\[0.1em]
                &\narrow{|}& \ident{float} \lsep \ident{decimal} \lsep \ident{int} \lsep \ident{bit} \lsep \ident{bool} \lsep \ident{string} 
 \\[0.6em] 
       \sigma &\narrow{=}& ~\hat{\sigma}~ \lsep \kvd{option}\langl \hat{\sigma} \rangl \lsep [\sigma] \\[0.1em]
              &\narrow{|}& \kvd{var}\langl\sigma_1, \ldots, \sigma_n\rangl \lsep ~\top~ \lsep \kvd{null}
\end{array}
\end{equation*}
%
The non-nullable types include records (consisting of an optional name and zero or more fields with
their types) and primitive types. Records arising from XML documents are always named, while records
used by JSON and CSV providers are always unnamed.

@@

Finally, a variant type \kvd{var} is annotated with the statically known types it may represent based
on the sample document. As discussed earlier (Section~\ref{sec:providers-xml}), the variant type forces
the user to handle the case when a value is of a different type than the ones statically known. For
this reason, the variant type also implicitly permits the \kvd{null} value.

@@

\begin{definition}
We write $\sigma_1 :> \sigma_2$ to denote that $\sigma_2$ is a subtype of $\sigma_1$. The 
subtyping relation is defined as a transitive reflexive closure of the following rules:

@@

\noindent\boxed{\emph{Variant types}}
\begin{align}
\label{eq:sub-var-bot}\tag{U1}
\kvd{var}\langl \sigma_1, \ldots, \sigma_n\rangl :>&~ \sigma \\
\label{eq:sub-var-eq}\tag{U2}
\kvd{var}\langl \sigma_1, \ldots, \sigma_n \rangl :>&~ \kvd{var}\langl \sigma_1', \ldots, \sigma_n' \rangl
\end{align}

\noindent\boxed{\emph{Record types}}
\begin{align}
\hspace{-1em}
\label{eq:sub-record1}\tag{R1}
 \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} \hspace{-0.1em}:> \hspace{-0.1em}
 \nu_?~\{ \nu_1\!:\!\sigma_1', .., \nu_n\!:\!\sigma_n' \}
 \; (\sigma_i \hspace{-0.25em}:>\hspace{-0.25em} \sigma_i')
\end{align}
\vspace{-1.4em}
\begin{align}
\hspace{-1em}
\label{eq:sub-record2}\tag{R2}
 \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} \hspace{-0.1em} :> \hspace{-0.1em}
 \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_m\!:\!\sigma_m \}
 \; (m \hspace{-0.1em}\geq\hspace{-0.1em} n)
\end{align}
\vspace{-1.0em}
\begin{align}
\label{eq:sub-record3}\tag{R3}
\hspace{-1.5em}
\begin{array}{l}
 \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \} :> \\
 \nu_?~\{ \nu_{\pi(1)}\!:\!\sigma_{\pi(1)}, .., \nu_{\pi(m)}\!:\!\sigma_{\pi(m)} \}
\end{array} \qquad &(\textnormal{permutation}~\pi)
\end{align}
\vspace{-0.6em}
\begin{align}
\label{eq:sub-record4}\tag{R4}
\hspace{-1.5em}
\begin{array}{l}
 \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n, \nu_{n+1}\!:\!\kvd{null}, .., \nu_{n+m}\!:\!\kvd{null} \} :> \\
 \nu_?~\{ \nu_1\!:\!\sigma_1, .., \nu_n\!:\!\sigma_n \}
\end{array}
\end{align}
\end{definition}

@@

\item The \ident{bit} type is a subtype of both \ident{int} and \ident{bool} (\ref{eq:sub-prim}). 
  Thus, a sample $0,1$ has a type \ident{bit}; $0,1,\kvd{true}$ is \ident{bool} and
  $0,1,2$ becomes \ident{int} (for a sample $0,1,2, \kvd{true}$ we would need a variant type).

@@

\item There is a top type (\ref{eq:sub-top}) and variants can be seen as the bottom because
  any type $\sigma$ is a subtype of a variant (\ref{eq:sub-var-bot}). There is no \emph{single} 
  bottom type -- there is a range of distinct variant types, but any variant is a subtype
  of any other variant (\ref{eq:sub-var-eq}).

@@

\noindent
If we were proposing a type system for type-checking programs, the use of multiple variant types
(which all are subtypes of each other) would not add any value and we could introduce a single 
bottom type instead. However, our purpose is different -- we aim to make reading structured data
easier and, as shown in Section~\ref{sec:providers-xml}, the types forming a variant provide a
useful information about the sample document.

Another important rule is (\ref{eq:sub-record4}), which allows subtype to have fewer record 
elements. In some cases, this allows us to infer records rather than variants. For example, given 
$\{ \ident{name}\!:\!\ident{string} \}$ and $\{ \ident{name}\!:\!\ident{string}, \ident{age}\!:\!\ident{int} \}$,
we can find a common supertype $\{ \ident{name}\!:\!\ident{string}, \ident{age}\!:\!\ident{int}~\kvd{option} \}$
which is also a record. This is more precise and more usable type than another common supertype
$\kvd{var}\langl \{ \ident{name}:\ident{string} \}, \{ \ident{name}:\ident{string}, \ident{age}:\ident{int} \}\rangl$.
The next section describes precisely how our inference algorithm works.

@@

We could also be more or less strict about missing data -- we choose to handle missing values 
silently when we can (\kvd{null} becomes an empty collection), but we are explicit in other cases 
(we infer \ident{string}~\kvd{option} as a hint to the user rather than treating \kvd{null} as an 
empty string). These choices are based on experience with using F\# Data in practice
and might differ for other languages.

@@

As mentioned earlier, the structural type inference relies on a finding common supertype. However,
the partially ordered set of types does not have a unique greatest lower bound. For example, record types
$\{\ident{a}\!:\!\ident{int}\}$ and $\{\ident{b}\!:\!\ident{bool}\}$ have common supertypes
$\{\ident{?a}\!:\!\ident{int}, \ident{?b}\!:\!\ident{bool}\}$ and 
$\kvd{var}\langl\{\ident{a}\!:\!\ident{int}\}, \{\ident{b}:\ident{bool}\}\rangl$

NOT TRUE
which are unrelated.

 Our inference algorithm always prefers records over unions (Theorem~\ref{thm:no-unions}). 
The definition of the \emph{common supertype} relation uses a number of auxiliary definitions 
that are explained in the discussion that follows.
